---
title: "Red Wine Exploratory Data Analysis"
author: "Shrey Marwaha"
date: "September 20, 2017"
output: html_document
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Loading required libraries
library(ggplot2)
library(stringr)
library(GGally)
library(reshape2)
library(dplyr)
library(tidyr)
library(gridExtra)
```

## Objectives

Through this project, I aim to explore and analyse the Red Wine Data Set, and infer the factors on which the quality of wine depends. I will start by exploring the data to get a feel of the information available and then, if possible, I hope to draw out the correlations between the various vaiables and Wine Quality. Through this, I aim to create a predictive model and check it against a test data.

```{r echo=FALSE, Load_the_Data}
# Load the Data
rdw <- read.csv('wineQualityReds.csv')

# Wrangling the data
rdw$quality <- factor(rdw$quality, ordered = TRUE)
#Creating new ordered categorical variable Class
rdw$class <- factor(ifelse(rdw$quality < 5, 'poor', 
                              ifelse(rdw$quality < 7, 'normal', 'exceptional')
                              ))

rdw$class <- ordered(rdw$class, levels = c('poor','normal','exceptional'))
```

## Structure and summary of the Dataframe

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
str(rdw)
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
summary(rdw)
```
# Univariate Plots Section

As a start, I see from the above statistics that there are a lot more number of **normal** red wines in the data, than 'poor' and 'exceptional'. Let's have a visual look at this by checking quality factor.

```{r echo=FALSE, Univariate_Plots}
ggplot(data = rdw, aes(x = quality))+
  geom_bar(fill = I('#40B0B5'))
```

That shows it quite clearly, as is mentioned in the Info file. I wonder if wines with quality rating 8 (exceptional) and quality rating 3 (poor) are outliers.

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Plotting all variables together, ommitting factor variables
rdw_compiled <- gather(rdw, "variable", "value", 2:12, factor_key = TRUE)
ggplot(aes(x = value), data = rdw_compiled)+
  geom_histogram(fill = I('#40B0B5') )+
  facet_wrap(~variable, scales = "free")
```

From the above initial plots, a lot of interesting things can be seen. Many variables are showing some kind of a relationship which can be of particular interest to us. Let's examine some of these further. I'll go about doing this category wise, namely: Acidity, Chemicals and Other features.

###Acidity

From the data, we can see that pH, fixed.acidity, volatile.acidity and citric.acid basically correspond to this category. Let's have a look at their distributions, together.

```{r echo=FALSE, message=FALSE, warning=FALSE}
p1 <- ggplot(data = rdw, aes(x = fixed.acidity))+
  geom_histogram(fill = '#338C90', binwidth = 0.1)+
  coord_cartesian(xlim = c(4, 14))
p2 <- ggplot(data = rdw, aes(x = volatile.acidity))+
  geom_histogram(fill = '#338C90', binwidth = 0.01)+
  coord_cartesian(xlim = c(0, 1))
p3 <- ggplot(data = rdw, aes(x = citric.acid))+
  geom_histogram(fill = '#338C90', binwidth = 0.01)
p4 <- ggplot(data = rdw, aes(x = pH))+
  geom_histogram(fill = '#338C90')
grid.arrange(p1, p2, p3, p4, ncol = 1)
```

From the above plots, we can see that fixed.acidity has more of a positively skewed distribution, it probably may be bimodal. The volatile.acidity variable also looks like a skewed distribution, though we may want to look at it further. There isn't much to be infered from the citric.acid plot, and the pH plot has a uniform distribution with most values concentrated between 3.0 and 3.5. That's quite accurate as wines are mildly acidic. The acidity adds to it's crispiness. 

I would like to see some x-axis transformations in citric.acid plot to extract any available information.

```{r echo=FALSE, message=FALSE}
ggplot(data = rdw, aes(x = citric.acid))+
  geom_histogram(fill = '#338C90', binwidth = 0.01)+
  scale_x_sqrt()
```

After a sqrt tranformation on x-axis, I can see that although there isn't relation visible, but many of the values have zero amount of citric acid in them. This probably depends on the flavor of the wine, and probably affects the quality.

###Chemicals
Let's check for presence of chemicals such as chlorides, sulphates & sulfur dioxide.

```{r echo=FALSE, message=FALSE}
c1 <- ggplot(data = rdw, aes(x = chlorides))+
  geom_histogram(fill = '#47979B')
c2 <- ggplot(data = rdw, aes(x = sulphates))+
  geom_histogram(fill = '#47979B')

grid.arrange(c1, c2, ncol = 1)
```

Both these curves show skewed distribution, and a large number of outliers. It's possible that these abnormal values may distinguish the quality/class of the wine, as presence of long tailed data is visible. 

```{r echo=FALSE, message=FALSE}
s1 <- ggplot(data = rdw, aes(x = free.sulfur.dioxide))+
  geom_histogram(fill = '#7EB6B9')+
  coord_cartesian(xlim = c(0, 60))
s2 <- ggplot(data = rdw, aes(x = total.sulfur.dioxide))+
  geom_histogram(fill = '#7EB6B9')+
  coord_cartesian(xlim = c(0, 200))

grid.arrange(s1, s2, heights = unit(1, 'npc'))
```

Both these variables show somewhat similar skewness. The free sulphur dioxide amount peaks at two places, first at 7 which is the median, and then around 9-10. Both have a small number of outliers that have been omitted from the visualizations.

###Other Parameters

Let's look at residual sugar, density and alcohol levels.

```{r echo=FALSE, message=FALSE}
# Residual Sugar 
o1 <- ggplot(data = rdw, aes(x = residual.sugar))+
  geom_histogram(fill = '#7eb6b9')
# Density
o2 <- ggplot(data = rdw, aes(x = density))+
  geom_histogram(fill = '#7e99b9')
#Alcohol
o3 <- ggplot(data = rdw, aes(x = alcohol))+
  geom_histogram(fill = '#7eb99f')
grid.arrange(o1, o2, o3, ncol = 1)
```

Residual sugar has very high outliers, while the 75th percentile is at 2.60 only. The distribution is positively skewed with median around 2.20. Density seems to be a uniform distribution, which can hence be a deciding factor in quality. Alcohol percentages are mostly positively skewed.

# Univariate Analysis

> **Tip**: Now that you've completed your univariate explorations, it's time to
reflect on and summarize what you've found. Use the questions below to help you
gather your observations and add your own if you have other thoughts!

### What is the structure of your dataset?

### What is/are the main feature(s) of interest in your dataset?

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

### Did you create any new variables from existing variables in the dataset?

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?


# Bivariate Plots Section

> **Tip**: Based on what you saw in the univariate plots, what relationships
between variables might be interesting to look at in this section? Don't limit
yourself to relationships between a main output feature and one of the
supporting variables. Try to look at relationships between supporting variables
as well.

```{r echo=FALSE, Bivariate_Plots}

```

# Bivariate Analysis

> **Tip**: As before, summarize what you found in your bivariate explorations
here. Use the questions below to guide your discussion.

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

### What was the strongest relationship you found?


# Multivariate Plots Section

> **Tip**: Now it's time to put everything together. Based on what you found in
the bivariate plots section, create a few multivariate plots to investigate
more complex interactions between variables. Make sure that the plots that you
create here are justified by the plots you explored in the previous section. If
you plan on creating any mathematical models, this is the section where you
will do that.

```{r echo=FALSE, Multivariate_Plots}

```

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

### Were there any interesting or surprising interactions between features?

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

------

# Final Plots and Summary

> **Tip**: You've done a lot of exploration and have built up an understanding
of the structure of and relationships between the variables in your dataset.
Here, you will select three plots from all of your previous exploration to
present here as a summary of some of your most interesting findings. Make sure
that you have refined your selected plots for good titling, axis labels (with
units), and good aesthetic choices (e.g. color, transparency). After each plot,
make sure you justify why you chose each plot by describing what it shows.

### Plot One
```{r echo=FALSE, Plot_One}

```

### Description One


### Plot Two
```{r echo=FALSE, Plot_Two}

```

### Description Two


### Plot Three
```{r echo=FALSE, Plot_Three}

```

### Description Three

------

# Reflection

> **Tip**: Here's the final step! Reflect on the exploration you performed and
the insights you found. What were some of the struggles that you went through?
What went well? What was surprising? Make sure you include an insight into
future work that could be done with the dataset.

> **Tip**: Don't forget to remove this, and the other **Tip** sections before
saving your final work and knitting the final report!